Top-down and bottom-up cohesiveness\\in microbial community coalescence

### 2021-06-03 draft/draft_v4.tex
draft v4 and fig4 v4 updated (pending update in overleaf)

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms,
\cite{Mansour2018,Luo2020,Vass2021}
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host's secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.

@@

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant community members has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasion success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
We refer to these two opposite scenarios as
``top-down'' or ``bottom-up'' forms of community cohesiveness, respectively.
Top-down cohesiveness emerges when the dominant invader co-selects other sub-dominant taxa
into the final community during coalescence.
Alternatively, bottom-up cohesiveness refers to the case when the dominant is co-selected by
the more rare members of its community.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Previous theoretical and computational studies suggest that the answer is determined by the type and
strength of the interactions of the community members with one another and with the environment
\cite{Tikhonov2016,Vila2019,Lechon2021},
but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

@@

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection in our experiments,
we found it useful to study a minimal model of community coalescence
(\methodsref{sim-min}).
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
respectively)
are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
They secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
off which
the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
that can in turn
be metabolized by the corresponding dominants.
With this structure, community members interact via simple cross-feeding networks that can be
very vertical if the dominant cross-feeds the sub-dominant but the sub-dominant does not cross-feed
the dominant (i.e. if the dominant cannot utilize the sub-dominant's secretions),
or more horizontal (bidirectional) if the cohort also cross-feeds the dominant.
We hypothesize that when two such communities coalesce,
bottom-up ecological co-selection can emerge if the invasive community is
sustained by a bidirectional cross-feeding network.

@@

Our simulations suggest that the strength and direction of ecological co-selection is modulated
by the underlying cross-feeding networks that shape the structure of
communities in synthetic minimal environments
\cite{Goldford2018,Estrela2020}.
This idea is supported by the observation that our Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
It also suggests that rarer taxa may play a more prominent role in co-selecting
dominant species when the metabolic fluxes across the community are
distributed rather than hierarchical.
These observations,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions of microbes with one another
and with the environment should be generically expected to produce ecological co-selection
during community coalescence.
Testing this theoretical prediction would require to map the cross-feeding networks of all of our communities.
Keeping track of every molecule secreted by every species in co-culture and by which species they are uptaken
is still a low throughput process that is both labor intensive and expensive.
Recent progress in metabolomic tools promise to help us test this hypothesis in future work.

@@

Our minimal model is set within the same MicroCRM framework that we used
for the previous simulations.
As described in the main text, the model contains two communities
of two species each
($\mathrm{s}_1$ to $\mathrm{s}_4$),
with five resources in total, out of which the first one ($\mathrm{R}_1$)
is replenished externally at the beginning of each growth cycle
and the rest correspond to the species' metabolic byproducts.
Each species secretes a unique byproduct, meaning that the metabolic
matrix $\mathbf{D}$ is binary in this case.
The specific structure of $\mathbf{D}$ is displayed below
--because it is a 3-dimensional matrix in our framework,
we have ``sliced'' it into the four 2-dimensional matrices corresponding to
our four species.

@@

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

@@

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_5$} \!\!\!
\end{matrix*}

@@

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_2$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_3$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_4$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_5$}
\end{matrix}

@@

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{s}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_4$} \!\!\!
\end{matrix*}

@@

\begin{pmatrix}
c_{11} & 0 & c_{13} & 0 & 0 \\
0 & 100 & 0 & 0 & 0 \\
c_{31} & 0 & 0 & 0 & c_{35} \\
0 & 0 & 0 & 100 & 0
\end{pmatrix}

\end{matrix}
\end{equation*}
%
Where we made the sub-dominants equally strong consumers of their dominants' secretions
($c_{22} = c_{44}$ = 100),
and we varied all other uptake rates depending on the scenario we were considering (see main text).
Whenever we were interested in the ratio between two rates (e.g. $c_{35}/c_{13}$ in \figref[e]{fig4})
we gave the one in the denominator a fixed value of 1 and let one in the numerator range
within the specified limits.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v3.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Illustration of the model structure and parameters.
The primary resource ($\mathrm{R}_1$) is replenished after
each growth-dilution cycle (red arrows).
Solid arrows indicate resource consumption, dashed arrows represent
resource secretion.
\textbf{b-e.} Coalescence outcomes in the minimal model
under different relations of cohesiveness between the
resident and the invasive communities.
We represent the relative Bray-Curtis similarity
between the invasive and the coalesced communities ($Q$) as a function of
the relevant model parameters.
For the specific representative cases indicated by the hollow circles,
we also show $Q$ as a function of the frequency of the invasive dominant
in pairwise competition with the resident dominant,
as well as the frequency of the invasive dominant invading alone versus invading
accompanied by its cohort.}
\label{fig4}
\end{figure}

>>>>

\textcolor{red}{To do: figSX, re-write minimal model, check literature, check p-values.}

@@

Microbial communities often invade one another.
This has been observed, for instance, 
in river courses where terrestrial microbes mix with aquatic microorganisms
\cite{Mansour2018,Luo2020,Vass2021},
or in soil communities being invaded as a result of
tillage and outplanting \cite{Rillig2016} or by
aerially dispersed bacteria and funghi \cite{Evans2019}.
Gut microbiomes can invade external communities
through the host's secretions \cite{Dutton2021},
and the skin microbiota is also subject to
invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.

@@

The mechanisms of ecological co-selection during community coalescence are still poorly understood.
Do a few key species recruit everyone else, or are collective interactions among all species
(including the rarer members of the community) relevant for coalescence outcomes?
While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
natural communities tend to be highly diverse \cite{Louca2016}
and
the role played by the less abundant community members has long been subject to debate \cite{Winfree2015}.
Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
The fate of these sub-dominant taxa may be dependent on the invasion success of their dominant
species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
or other forms of facilitation from the rarer members of the population.
We refer to these two opposite scenarios as the
``top-down'' or ``bottom-up'' forms of community cohesiveness, respectively.
Top-down cohesiveness emerges when the dominant invader co-selects other sub-dominant taxa
into the final community during coalescence.
Alternatively, bottom-up cohesiveness refers to the case when the dominant is co-selected by
the more rare members of its community.
Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
as illustrated in \figref[e]{fig1}.
Which of these situations are typically found in nature?
Previous theoretical and computational studies suggest that the answer is determined by the type and
strength of the interactions of the community members with one another and with the environment
\cite{Tikhonov2016,Vila2019,Lechon2021},
but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.

@@

To better understand the underlying mechanisms that govern
the emergence of ecological co-selection in our experiments,
we found it useful to study a minimal model of community coalescence
(\methodsref{sim-min}).
This model is comprised of two communities with two species each
as illustrated in \figref[a]{fig4}.
Within each community, the dominant species
($\mathrm{S}_1$ and $\mathrm{s}_1$ in the resident and invasive communities
respectively)
are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
They secrete a single byproduct
($\mathrm{R}_2$ and $\mathrm{r}_2$ respectively)
off which
the sub-dominants ($\mathrm{S}_2$ and $\mathrm{s}_2$ respectively) can feed.
Finally, these sub-dominants secrete an additional resource
($\mathrm{R}_3$ and $\mathrm{r}_3$ respectively)
that can in turn
be metabolized by the corresponding dominants.
With this structure, community members interact via simple cross-feeding networks that can be
very vertical if the dominant cross-feeds the sub-dominant but the sub-dominant does not cross-feed
the dominant (i.e. if the dominant cannot utilize the sub-dominant's secretions),
or more horizontal (bidirectional) if the cohort also cross-feeds the dominant.
We hypothesize that when two such communities coalesce,
bottom-up ecological co-selection can emerge if the invasive community is
sustained by a bidirectional cross-feeding network.

@@

Our simulations suggest that the strength and direction of ecological co-selection is modulated
by the underlying cross-feeding networks that shape the structure of
communities in synthetic minimal environments
\cite{Goldford2018,Estrela2020}.
This idea is supported by the observation that our Microbial Consumer-Resource Model
captures the trends observed experimentally when we enable a large variation
in the metabolic fluxes across species.
The model predicts a trade-off between the strength of bottom-up co-selection
and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
which we have confirmed experimentally.
It also suggests that rarer taxa may play a more prominent role in co-selecting
dominant species when the cross-feeding interactions across community members are
horizontal rather than hierarchical.
Testing this theoretical prediction would require to map the cross-feeding networks of all of our communities.
Keeping track of every molecule secreted by every species in co-culture and by which species they are uptaken
is still a low throughput process that is both labor intensive and expensive.
Recent progress in metabolomic tools promise to help us test this hypothesis in future work.
Our findings,
together with previous results in different systems \cite{Sierocinski2017} as well as
theoretical predictions
\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
suggest that collective interactions of microbes with one another
and with the environment should be generically expected to produce ecological co-selection
during community coalescence.

@@

Our minimal model is set within the same MicroCRM framework that we used
for the previous simulations.
As described in the main text, the model contains two communities
of two species each
($\mathrm{S}_1$ and $\mathrm{S}_2$ in the resident community,
$\mathrm{s}_1$ and $\mathrm{s}_2$ in the invasive community),
with five resources in total, out of which the first one ($\mathrm{R}_1$)
is replenished externally at the beginning of each growth cycle
and the rest correspond to the species' metabolic byproducts.
Each species secretes a unique byproduct, meaning that the metabolic
matrix $\mathbf{D}$ is binary in this case.
The specific structure of $\mathbf{D}$ is displayed below
--because it is a 3-dimensional matrix in our framework,
we have ``sliced'' it into the four 2-dimensional matrices corresponding to
our four species.

@@

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
\text{\scriptsize $\mathrm{r}_2$} \!\!\! & 
\text{\scriptsize $\mathrm{r}_3$}
\end{matrix}

@@

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
\text{\scriptsize $\mathrm{r}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{r}_3$} \!\!\!
\end{matrix*}

@@

\begin{matrix}
\text{\scriptsize $\mathrm{R}_1$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_2$} \;\;\; & 
\text{\scriptsize $\mathrm{R}_3$} \;\;\; & 
\text{\scriptsize $\mathrm{r}_2$} \;\;\; & 
\text{\scriptsize $\mathrm{r}_3$}
\end{matrix}

@@

\begin{matrix*}[l]
\text{\scriptsize $\mathrm{S}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{S}_2$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_1$} \!\!\! \\
\text{\scriptsize $\mathrm{s}_2$} \!\!\!
\end{matrix*}

@@

\begin{pmatrix}
C_{11} & 0 & C_{13} & 0 & 0 \\
0 & 100 & 0 & 0 & 0 \\
c_{11} & 0 & 0 & 0 & c_{13} \\
0 & 0 & 0 & 100 & 0
\end{pmatrix}

\end{matrix}
\end{equation*}
%
Where we made the sub-dominants equally strong consumers of their dominants' secretions
($C_{22} = c_{22}$ = 100),
and we varied all other uptake rates depending on the scenario we were considering (see main text).
Whenever we were interested in the ratio between two rates (e.g. $c_{13}/C_{13}$ in \figref[e]{fig4})
we gave the one in the denominator a fixed value of 1 and let one in the numerator range
within the specified limits.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v4.pdf}
\caption{\textbf{A minimal model of community coalescence.}
\textbf{a.} Illustration of the model structure and parameters.
The primary resource ($\mathrm{R}_1$) is replenished after
each growth-dilution cycle (red arrows).
Solid arrows indicate resource consumption, dashed arrows represent
resource secretion.
\textbf{b-e.} Coalescence outcomes in the minimal model
under different relations of cohesiveness between the
resident and the invasive communities.
We represent the relative Bray-Curtis similarity
between the invasive and the coalesced communities ($Q$) as a function of
the relevant model parameters.
For the specific representative cases indicated by the hollow circles,
we also show $Q$ as a function of the frequency of the invasive dominant
in pairwise competition with the resident dominant,
as well as the frequency of the invasive dominant invading alone versus invading
accompanied by its cohort.}
\label{fig4}
\end{figure}
